Classical Dynamics of Harmonically Trapped Interacting Particles

### 2017-12-19 draft_09.tex
added many citations

  Motivated by the interest in the dynamics of trapped quantum gases, we study the microcanonical
  dynamics of a trapped gas of classical particles interacting with a finite-range interaction of
  tunable strength.  We examine two questions which have been of interest in quantum dynamics: (1)
  the breathing mode (size oscillation) dynamics of the trapped gas and the dependence of the
  breathing frequency on the interaction strength, and (2) the long-time relaxation and possible
  thermalization of the finite isolated gas.

@@

The non-equilibrium dynamics of isolated systems has been the subject of a large volume of work in
recent years.  In particular, in the quantum context, the study of unitary quantum evolution of
many-body and few-body systems has undergone an explosive growth.  This interest was partly fueled
by the availability of experimental systems, e.g., trapped ultracold atomic and ionic systems, where
non-dissipative dynamics can be studied explicitly due to excellent isolation.  There are also
foundational reasons for studying quantum dynamics in isolation: recent activity has led to advances
in the understanding of basic principles connecting quantum dynamics, quantum chaos and the
emergence of statistical mechanics and thermalization (CITE).  For classical many-body systems,
explorations of the connections between microscopic dynamical rules, chaos, ergodicity and
statistical mechanics have a much longer history, dating back to the time of Boltzmann and Poincare
\cite{Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book, EckmannRuelle_RMP85}.

@@

Experiments with cold atoms, which have an essential motivational role for the current interest in
many-body quantum dynamics, are generally performed in a trapping potential, either optical or
magnetic.  As a result, in the last two decades many questions of many-body quantum physics have
been reformulated in the presence of a harmonic trap.  The presence of a harmonic trap often results
in new dynamical phenomena.  For example, the motion of vortices in a trapped condensate (CITE) has
new features, such as precession, compared to vortex motion in uniform condensates (CITE).  Trapping
also results in new collective phenomena which have no analog in uniform situations: for example,
collective modes like dipole, breathing and quadrupole modes are specific to trapped many-body
systems.  Dipole modes (center of mass oscillations) and breathing modes (size oscillations) appear
even for one-dimensional trapped systems; higher-dimensional systems display additional, more
complicated modes like quadrupole and scissors modes.  Because of the omnipresence of trapping
potentials in cold-atom experiments, dipole and breathing mode dynamics are pervasive across the
field: These modes have been intensively studied and used for diagnostic purposes since the early
days of research with trapped quantum-degenerate gases, and continue to generate interest today
\cite{Naegerl_Science2009, Esslinger_exp_PRL2003, collective_modes_early_papers,
  Tschischik_BHbreathing_PRA_2013, KroenkeSchmelcher_BM, Bonitz_Review, Bouchoule_PRL2014,
  our_superTG_PRA_2015, STG_Astrakharchik_PRL2005, STG_Chen-etal_PRA2010,
  br_mode_long_range_interactions, 1D_breathing_mode_recent, KohnMode, Cominotti_2015,
  ZarembaTowers_PRA2014, Wetterich_JPB2011, fermion2Dbreathing, Stringari_lowDcollective_PRA15,
  Mazets_1Dbreathing_EPJD11}.  (CITE).
%
In one of the best-known experiments on quantum many-body dynamics,
Ref.~\cite{NewtonsCradle_Nature06}, the dynamics of an integrable system (Lieb-Liniger bosons) was
seen to have ultraslow relaxation in a trap, supposedly due to integrability.  This experiment has
led to continuing efforts to understand how a trap breaks integrability (CITE).

@@

Our first theme will be the breathing mode of the classical interacting gas, and the effect of
interactions on the frequency of the breathing mode.  In the quantum case, this has long been a
topic of interest for both bosons and fermions.  For bosons with short-range (contact) interactions,
the situation has been studied in one, two and three dimensions.  At zero interactions, the
breathing mode is twice the trapping frequency $\omega_0$ in each case, assuming isotropic trapping.
A mean field (Gross-Pitaevskii) calculation gives the frequency to be $\sqrt{D+2}$ times the
trapping frequency, where $D$ is the spatial dimension.  For three-dimensional (3D) traps, the
mean-field description can be expected to be qualitatively correct, thus the breathing mode
frequency is expected to change monotonically from $2\omega_0$ at zero interactions to
$\sqrt{5}\omega_0$ at large interactions.  For 2D traps, the breathing frequency is exactly
$2\omega_0$ for any interaction strength \cite{PitaevskiiRosch_PRA97}, due to a symmetry.  Classical
trapped particles with scale-invariant interaction potentials also have breathing frequency
$2\omega_0$ independent of interaction strength \cite{PitaevskiiRosch_PRA97}.  The interaction
dependence in 1D (trapped Lieb-Liniger gas) is also remarkable: as the interaction is increased, the
breathing frequency first decreases from $2\omega_0$ to the mean-field prediction
$\sqrt{3}\omega_0$, and then at larger interactions increases again, returning to $2\omega_0$ in the
infinite interaction (Tonks-Girardeau) limit \cite{Naegerl_Science2009,
  Tschischik_BHbreathing_PRA_2013, 1D_breathing_mode_recent, Bouchoule_PRL2014,
  KroenkeSchmelcher_BM}.
%
The breathing mode has also been investigated extensively for fermionic systems with contact
interactions \cite{GrimmSmith_unitaryfermions_PRA08, fermion2Dbreathing,
  Stringari_lowDcollective_PRA15}, and also for fermionic and bosonic systems with long-range
interactions \cite{br_mode_long_range_interactions, Bonitz_Review}.  The breathing mode has also
been studied for a high-temperature trapped gas where a Boltzmann-equation description is valid,
both theoretically \cite{Guery-Odelin1999} and experimentally
\cite{Cornell_classsical_3Dexpt_NatPhys15}.

@@

Our second theme is the process of relaxation at longer timescales, and possible thermalization.
This topic is also motivated by intense recent research in the quantum context.  In quantum isolated
systems, thermalization is now understood in terms of the so-called eigenstate thermalization
hypothesis (ETH), which postulates that in ergodic systems the expectation values of observables in
individual eigenstates depend ony on the eigenenergy, and hence represents the thermal value (CITE).
The connection to statistical mechanics is particularly difficult for explicitly finite systems, in
which case ETH is to be understood in terms of finite size scaling (CITE).  In classical systems,
the connection between microscopic dynamics and statistical mechanics has been continually studied
since the 19th century \cite{Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book,
  EckmannRuelle_RMP85}.  It is generally understood that systems with nonzero Lyapunov exponents,
i.e., chaotic or ergodic systems, thermalize in the long-time limit, provided there are enough
degrees of freedom.  As in the quantum case, finite systems are particularly intricate.  In finite
systems, defining thermalization is trickier as most observables will fluctuate or oscillate
substantially.  In addition, the influence of Kolmogorov-Arnold-Moser (KAM) orbits may play a role
in preventing thermalization; it is generally believed that the fraction of phase space where KAM
physics might be relevant decreases rapidly with particle number \cite{Dumas_book_KAMstory}.

@@

There is a significant literature on calculating Lyapunov exponents in model statistical-mechanics
systems and thus exploring ergodicity \cite{Dorfman_book_1999} (CITE).  This provides intuition of
what types of microscopic interactions are likely to produce thermalizability, and is an important
step in the program of understanding statistical mechanics ``from the bottom up''.  However, we are
not aware of a significant literature on following explicitly the microcanonical dynamics of
statistical-mechanical or few-body systems.  Several such studies have appeared recently (CITE),
motivated by the quantum dynamics literature, like the present work.  A few earlier studies have
also examined thermalization dynamics in one dimensional gravitational systems \cite{Yawn1997,
  Tsuchiya2000}.
%
Investigating such real-time dynamics can be expected to provide insights on the timescales and
microscopic mechanisms involved in the emergence of statistical mechanics from collections of
particles.

@@

The gas contains $N$ such identical particles, each of mass $m$, in a harmonic trap.  The
Hamiltonian describing the gas is 
\begin{equation}
  H ~=~
  \frac{1}{2}m\omega_0^2\sum_{i}x_i^2
  ~+~ \frac{1}{2}m\sum_i  v_i^2
  ~+~ \sum_{\left|x_i-x_j\right|\textless\sigma}F_0\left(\sigma-\left|x_i-x_j\right|\right)
\label{eq:preparation2}
\end{equation}
%
It is useful to rescale the quantities.  We will measure distance, time, energy and force in units
of $\sigma$, $1/\omega_0$, $m\omega_0^2\sigma^2$ and $m\omega_0^2\sigma$ respectively:
\begin{equation} \label{eq:transform}
  \tilde{x_i} = \frac{x_i}{\sigma}\ , \quad 
  \tilde{H} = \frac{H}{m\omega_0^2\sigma^2}\ ,  \quad \tilde{t} = \omega_0t \ ,  \quad 
\tilde{F_0} = \frac{F_0}{m\omega_0^2\sigma}\ .
\end{equation}
%
Eq.~\ref{eq:preparation2} is then rewritten as
\begin{equation}
  \tilde{H} =
  \frac{1}{2}\sum_{i}\tilde{x_i}^2
  + \frac{1}{2}\sum_i\left(\frac{d\tilde{x}_i}{d\tilde{t}}\right)^2 
  + \sum_{\left|\tilde{x_i}-\tilde{x_j}\right|\textless
    1}\tilde{F}_0\left(1-\left|\tilde{x_i}-\tilde{x_j}\right|\right)  
\end{equation}
%
Through this rescaling, we have reduced the number of the parameters in our model to three: the
energy $\tilde{H}$, the interaction strength $\tilde{F_0}$, and the number of particles $N$.  The
rescaling is equivalent to setting $\sigma$, $\omega_0$ and $m$ to $1$; this is what we do in our
numerical simulation.  In the rest of this paper, we will use ``$E$" and ``$F_0$" to denote the
reduced versions $\tilde{H}$ and $\tilde{F_0}$.

We will be concerned with the size of the cloud.  We quantify the size through the root-mean-square
of particle positions $\left\lbrace x_i\right\rbrace$, 
\begin{equation}
R\equiv\left(\overline{x_i^2}\right)^\frac{1}{2} \, ,
\label{eq:def_of_R}
\end{equation}
and refer to this quantity as the `radius' or `cloud radius'.
%
In the dilute gas limit, there are few or no interactions occuring at most times.  Thus the energy
$E$ is dominated by the trap energy, i.e., $E\approx m\omega_0^2\sum_{i}x_i^2= Nm\omega_0^2R^2$, or,
in our units,
\begin{equation}
  E\approx NR^2 \, .
  \label{eq:E_and_R}
\end{equation}
%
This approximation is good when the average distance betwen particles is much larger than the
interaction range $\sigma=1$, i.e., whenever $R\gg1$, which is the regime we consider dynamics in.
Another way of describing this dilute-gas regime is that the gas is far from the ground state
(described in the next subsection).  Neglecting the interaction energy to estimate $E$ is reasonable
even at very large interactions ($F_0\gg E/N$), because for very large interactions, the particles
behave as hard rods which bounce off each other very rapidly, so that at most instants one can
expect no collisions to be taking place, provided the gas is dilute.  It is also expected to be a
good approximation at very small $F_0$ since we can then simply neglect the interaction energy.
Thus, we will use this as a common estimate for $R$ in terms of the energy.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbhp]
\centering
\includegraphics[scale=0.6]{ZhiyuPictures/N=5_GS_pre_2_rev.eps}
\caption{The ground state of a system with $N=5$ particles, shown via the positions of each
  particle, labeled 1 through 5.  At large interactions, the particles position themselves just
  outside the range of interactions of the neighboring particles.}
\label{fig:GS1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

Because of the `Heaviside theta' form of our interaction, at large enough interactions (large $F_0$)
the particles are spaced exactly at distance equal to the range $\sigma$, which is distance $1$ in
our rescaled units.  The interaction is then effectively a `hard-core' interaction.

@@

The initial state is taken to have particles with zero velocity and random positions; hence the line
distribution in the phase space picture.  This has the advantage that the breathing motion is
prominently visible.  In addition, the question of long-time relaxation has the simple
interpretation of evolving from the line distribution to a circularly symmetric distribution.

@@

We are interested in the oscillations of the size $R(t)$ of the cloud.  It is convenient that our
finite-size simulations start with a line distribution in phase space.  If we started with a state
whose configuration in phase-space deviates only slightly from circular symmetry, the oscillating
amplitude would be too small to be distinguished from noise.

@@

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h
]
\subfigure[]{
\begin{minipage}[b]{0.4\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/bounce.png}
\end{minipage}
}
\subfigure[]{
\begin{minipage}[b]{0.4\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/pass.png}
\end{minipage}
}
\caption{Schematic diagrams of two-particle collision process in the $X_{\rm rot}$-$P_{\rm rot}$
  plane. \textbf{(a)}When $F_0$ is large, particles cannot pass each other.  Being subject to a
  constant repulsive force, their trajectories during collision are circular arcs.  For very strong
  interactions (red dashed line, straight), they exchange momenta instantaneously.  The yellow,
  green, and blue dashed curves are for successively weaker interactions.  \textbf{(b)} When $F_0$
  is small, particles pass each other, so that the force they feel is not continuous.  Their
  trajectory has a turning point. The dashed bright blue trajectory is identical with the one in
  (a), which is the critical situation between passing and bouncing. In that case, particles have
  zero relative velocity when they collide. }
\label{fig:Breathingfrequency3}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

To analyze the rotation relative to the $X_{\rm rot}$-$P_{\rm rot}$ frame, we consider two-particle
collisions.  In Figure \ref{fig:Breathingfrequency3}(a), we show schematics of such a collision near
the trap center at strong interactions.  For very large $F_0$, the particles simply exchange
momentum when they are at distance $\sigma$ from each other; this process is shown in red as two
straight lines.  (The particles are initially at $X=\pm\frac{1}{2}\sigma$ and remain at these $X$
values, but exchange their momenta.)  For smaller $F_0$, there is change of both position and
momentum, as shown in yellow, green and blue for successively weaker interactions. The blue line
shows a case where $F_0$ is smaller enough such that the particles can actually cross each other.

In Figure \ref{fig:Breathingfrequency3}(a) we show collisions for small $F_0$, so that the particles
pass each other.  The force changes direction discontinuously when the particles cross.  In the
$X_{\rm rot}$-$P_{\rm rot}$ plane, this is seen as a sharp turning point in the trajectory.

@@

The precession angle is $\alpha\sim \sigma/P$.  In our rescaled units, the typical momenta of
interacting partilces is of the order $R$.  (In phase space $R$ can be interpreted either as the
extent in real space when the velocities are small, or as the extent in momentum space when the
particles approach $x=0$, i.e. when the distribution is along the $P$ direction.)  Thus, we can
estimate $\alpha\sim \sigma/R$, i.e., $\alpha\sim 1/R$, since our unit of length is $\sigma$.  This
is the rotation per unit collision.  Since there are $\sim N$ collistions per period $T=2\pi$, we
have the estimate for the interaction-induced rotation per unit time as
\begin{equation}
\delta ~=~ 2 \frac{N}{2\pi} \frac{1}{R} ~=~  \frac{1}{\pi}  N^{3/2}E^{-1/2}
\label{eq:breathingfrequency1}
\end{equation}
The factor $2$ accounts for the fact that each rotation of the elongated cloud in phase space
corresponds to two breathing mode periods.  In the last step we have used the estimate $E\sim
NR^2$.

>>>>

Motivated by the interest in the dynamics of trapped quantum gases, we study the microcanonical
dynamics of a trapped one-dimensional gas of classical particles interacting via a finite-range
repulsive force of tunable strength.  We examine two questions which have been of interest in
quantum dynamics: (1) the breathing mode (size oscillation) dynamics of the trapped gas and the
dependence of the breathing frequency on the interaction strength, and (2) the long-time relaxation
and possible thermalization of the finite isolated gas.
%
We show that the breathing mode frequency has nonmonotonic dependence on the magnitude of the mutual
repulsion, decreasing for small interactions and increasing for larger interactions.  We explain
these dependences in terms of slowing-down or speeding-up effects of two-body collision processes.
We find that the gas thermalizes within a reasonable finite timescale in the sense of
single-particle energies acquiring a Boltzmann distribution, only when the interaction strength is
large compared to the energy per particle.

@@

The non-equilibrium dynamics of isolated systems has been the subject of a large volume of work in
recent years.  In particular, in the quantum context, the study of unitary quantum evolution of
many-body and few-body systems has undergone an explosive growth.  This interest was partly fueled
by the availability of experimental systems, e.g., trapped ultracold atomic and ionic systems, where
non-dissipative dynamics can be studied explicitly due to excellent isolation from the environment.
There are also foundational reasons for studying quantum dynamics in isolation: recent activity has
led to advances in the understanding of basic principles connecting quantum dynamics, quantum chaos
and the emergence of statistical mechanics and thermalization \cite{PolkovnikovRigol_AdvPhys2016,
  BorgonoviIzrailevSantos_PhysRep2016}.  For classical many-body systems, explorations of the
connections between microscopic dynamical rules, chaos, ergodicity and statistical mechanics have a
much longer history, dating back to the time of Boltzmann and Poincare \cite{Gaspard_book_1998,
  Dorfman_book_1999, Dumas_book_KAMstory, Boltzmann_legacy_book, EckmannRuelle_RMP85,
  Gaspard_PhysicaA06}.

@@

Experiments with cold atoms, which have an essential motivational role for the current interest in
many-body quantum dynamics, are generally performed in a trapping potential, either optical or
magnetic.  As a result, in the last two decades many questions of many-body quantum physics have
been reformulated in the presence of a harmonic trap.  The presence of a harmonic trap often results
in new dynamical phenomena.  For example, the motion of vortices in a trapped condensate
\cite{vortices_trappedBEC} has new features, such as precession, compared to vortex motion in
uniform condensates \cite{vortices_uniformBEC}.  Trapping also results in new collective phenomena
which have no analog in uniform situations: for example, collective modes like dipole, breathing and
quadrupole modes are specific to trapped many-body systems.  Dipole modes (center of mass
oscillations) and breathing modes (size oscillations) appear even for one-dimensional trapped
systems; higher-dimensional systems display additional, more complicated modes like quadrupole and
scissors modes.  Because of the omnipresence of trapping potentials in cold-atom experiments, dipole
and breathing mode dynamics are pervasive across the field: These modes have been intensively
studied and used for diagnostic purposes since the early days of research with trapped
quantum-degenerate gases, and continue to generate interest today
\cite{collective_modes_early_papers, PitaevskiiRosch_PRA97, Esslinger_exp_PRL2003, STG_Astrakharchik_PRL2005, 
  GrimmSmith_unitaryfermions_PRA08, OrignacCitro_PRA08, Naegerl_Science09, Wetterich_JPB2011, Mazets_1Dbreathing_EPJD11,
  fermion2Dbreathing, Tschischik_BHbreathing_PRA_2013, KroenkeSchmelcher_BM, Bouchoule_PRL2014,
  br_mode_long_range_interactions, Bonitz_Review, QuinnHaque_PRA14, Tschischik_BH_2papers,
   1D_breathing_mode_recent, Minguzzi_dipole_PRA2015, 
  Stringari_lowDcollective_PRA15,    MistakidisSchmelcher_PRA17}. 
%
In one of the best-known experiments on quantum many-body dynamics,
Ref.~\cite{NewtonsCradle_Nature06}, the dynamics of an integrable system (Lieb-Liniger bosons) was
seen to have ultraslow relaxation in a trap, supposedly due to integrability.  This experiment has
led to continuing efforts to understand how a trap breaks integrability
\cite{CauxKonik_glimmers_PRX15, CauxDoyonDubailKonik_1711, DoyonSpohn_JSTAT17, Moore_arXiv17.10}.

@@

Our first theme will be the breathing mode of the classical interacting gas, and the effect of
interactions on the frequency of the breathing mode.  In the quantum case, this has long been a
topic of interest for both bosons and fermions.  For bosons with short-range (contact) interactions,
the situation has been studied in one, two and three dimensions.  At zero interactions, the
breathing mode is twice the trapping frequency $\omega_0$ in each case, assuming isotropic trapping.
A mean field (Gross-Pitaevskii) calculation gives the frequency to be $\sqrt{D+2}$ times the
trapping frequency, where $D$ is the spatial dimension.  For three-dimensional (3D) traps, the
mean-field description is expected to be qualitatively correct, thus the breathing mode frequency is
expected to change monotonically from $2\omega_0$ at zero interactions to $\sqrt{5}\omega_0$ at
large interactions.  For 2D traps, the breathing frequency is exactly $2\omega_0$ for any
interaction strength \cite{PitaevskiiRosch_PRA97}, due to a symmetry.  Classical trapped particles
with scale-invariant interaction potentials also have breathing frequency $2\omega_0$ independent of
interaction strength \cite{PitaevskiiRosch_PRA97}.  The interaction dependence in 1D (trapped
Lieb-Liniger gas) is also remarkable: as the interaction is increased, the breathing frequency first
decreases from $2\omega_0$ to the mean-field prediction $\sqrt{3}\omega_0$, and then at larger
interactions increases again, returning to $2\omega_0$ in the infinite interaction (Tonks-Girardeau)
limit \cite{Naegerl_Science09, Tschischik_BHbreathing_PRA_2013, 1D_breathing_mode_recent,
  Bouchoule_PRL2014, KroenkeSchmelcher_BM}.
%
The breathing mode has also been investigated extensively for fermionic systems with contact
interactions \cite{GrimmSmith_unitaryfermions_PRA08, fermion2Dbreathing,
  Stringari_lowDcollective_PRA15}, and also for fermionic and bosonic systems with long-range
interactions \cite{OrignacCitro_PRA08, br_mode_long_range_interactions, Bonitz_Review}.  The
breathing mode has also been studied for a high-temperature trapped gas where a Boltzmann-equation
description is valid, both theoretically \cite{Guery-Odelin1999} and experimentally
\cite{Cornell_classsical_3Dexpt_NatPhys15}.

@@

Our second theme is the process of relaxation at longer timescales, and possible thermalization.
This topic is also motivated by intense recent research in the quantum context.  In quantum isolated
systems, thermalization is now understood in terms of the so-called eigenstate thermalization
hypothesis (ETH), which postulates that in ergodic systems the expectation values of observables in
individual eigenstates depend ony on the eigenenergy, and hence represent the thermal value
\cite{PolkovnikovRigol_AdvPhys2016, BorgonoviIzrailevSantos_PhysRep2016,
  ETH_Deutsch_Srednicki_Rigol}.  The connection to statistical mechanics is particularly difficult
for explicitly finite systems, in which case ETH is to be understood in terms of finite size scaling
\cite{Beugeling_ETHscaling_PRE14}.  In classical systems, the connection between microscopic dynamics and statistical mechanics
has been continually studied since the 19th century \cite{Dorfman_book_1999, Dumas_book_KAMstory,
  Boltzmann_legacy_book, EckmannRuelle_RMP85}.  It is generally understood that systems with nonzero
Lyapunov exponents, i.e., chaotic or ergodic systems, thermalize in the long-time limit, provided
there are enough degrees of freedom.  As in the quantum case, finite systems are particularly
intricate.  In finite systems, defining thermalization is trickier as most observables will
fluctuate or oscillate substantially.  In addition, the influence of Kolmogorov-Arnold-Moser (KAM)
orbits may play a role in preventing thermalization; it is generally believed that the fraction of
phase space where KAM physics might be relevant decreases rapidly with particle number
\cite{Dumas_book_KAMstory}.

@@

There is a significant literature on calculating Lyapunov exponents in model statistical-mechanics
systems and thus exploring ergodicity \cite{Dorfman_book_1999, deWijn_fine,
  Lyapunov_numerical_calculations, Lyapunov_analytical_calculations, FiniteTimeLyapunov}.  This
provides intuition of what types of microscopic interactions are likely to produce thermalizability,
and is an important step in the program of understanding statistical mechanics ``from the bottom
up''.  However, we are not aware of a significant literature on following explicitly the
microcanonical dynamics of statistical-mechanical or few-body systems.  Several such studies have
appeared recently \cite{DoyonSpohn_JSTAT17, Moore_arXiv17.10, JinKatsnelson_NJP13}, motivated by the
quantum dynamics literature, like the present work.  A few earlier studies have also examined
thermalization dynamics in one dimensional gravitational systems \cite{Yawn1997, Tsuchiya2000}.
%
Investigating such real-time dynamics can be expected to provide insights on the timescales and
microscopic mechanisms involved in the emergence of statistical mechanics from collections of
particles.

@@

The gas contains $N$ such identical particles, each of mass $m$, in a harmonic trap.  The
Hamiltonian describing the gas is 
\begin{equation}
  H ~=~
  \frac{1}{2}m\omega_0^2\sum_{i}x_i^2
  ~+~ \frac{1}{2}m\sum_i  v_i^2
  ~+~ \sum_{\left|x_i-x_j\right|\textless\sigma}F_0\left(\sigma-\left|x_i-x_j\right|\right)
\label{eq:preparation2}
\end{equation}
%
Here $F_0\left(\sigma-\left|r\right|\right)\theta(\sigma-|r|)$, with $r=x_i-x_j$ and $\theta()$ the
Heaviside theta function, is the potential corresponding to the force introduced in
Eq.\ \eqref{eq:preparation1}.
%
It is useful to rescale the quantities.  We will measure distance, time, energy and force in units
of $\sigma$, $1/\omega_0$, $m\omega_0^2\sigma^2$ and $m\omega_0^2\sigma$ respectively:
\begin{equation} \label{eq:transform}
  \tilde{x_i} = \frac{x_i}{\sigma}\ , \quad 
  \tilde{H} = \frac{H}{m\omega_0^2\sigma^2}\ ,  \quad \tilde{t} = \omega_0t \ ,  \quad 
\tilde{F_0} = \frac{F_0}{m\omega_0^2\sigma}\ .
\end{equation}
%
Eq.~\ref{eq:preparation2} is then rewritten as
\begin{equation}
  \tilde{H} =
  \frac{1}{2}\sum_{i}\tilde{x_i}^2
  + \frac{1}{2}\sum_i\left(\frac{d\tilde{x}_i}{d\tilde{t}}\right)^2 
  + \sum_{\left|\tilde{x_i}-\tilde{x_j}\right|\textless
    1}\tilde{F}_0\left(1-\left|\tilde{x_i}-\tilde{x_j}\right|\right)  
\end{equation}
%
Through this rescaling, we have reduced the number of the parameters in our model to three: the
energy $\tilde{H}$, the interaction strength $\tilde{F_0}$, and the number of particles $N$.  The
rescaling is equivalent to setting $\sigma$, $\omega_0$ and $m$ to $1$; this is what we do in our
numerical simulation.  In the rest of this paper, we will use ``$E$" and ``$F_0$" to denote the
reduced versions $\tilde{H}$ and $\tilde{F_0}$, and omit the tilde when writing rescaled
quantities.

In the limit of infinite $F_0$, the particles maintain distance larger than $\sigma$ from each
other.  In this limit each particle can then be thought of as a `hard rod' of length $2\sigma$.  In
the absence of a trap, this would be the one-dimensional classical system shown by Tonks to be
classically integrable \cite{Tonks_1936}.

\subsection{Cloud size or `radius'}

We will be concerned with the size of the cloud.  We quantify the size through the root-mean-square
of particle positions $\left\lbrace x_i\right\rbrace$, 
\begin{equation}
R ~=~ \left(\overline{x^2}\right)^\frac{1}{2} ~=~ \left(\frac{1}{N}\sum_i{x_i^2}\right)^\frac{1}{2} \, ,
\label{eq:def_of_R}
\end{equation}
and refer to this quantity as the `radius' or `cloud radius'.
%
In the dilute gas limit, there are few or no interactions occuring at most times.  Thus the energy
$E$ is dominated by the trap energy, i.e., $E\approx m\omega_0^2\sum_{i}x_i^2= Nm\omega_0^2R^2$, or,
in our units,
\begin{equation}
  E\approx NR^2 \, .
  \label{eq:E_and_R}
\end{equation}
%
This approximation is good when the average distance betwen particles is much larger than the
interaction range $\sigma=1$, i.e., whenever $R\gg1$, which is the regime we consider dynamics in.
Another way of describing this dilute-gas regime is that the gas is far from the ground state
(described in the next subsection).  Neglecting the interaction energy to estimate $E$ is reasonable
even at very large interactions ($F_0\gg E/N$), because for very large interactions, the particles
behave as hard rods which bounce off each other very rapidly, so that at most instants one can
expect no collisions to be taking place, provided the gas is dilute.  It is also expected to be a
good approximation at very small $F_0$ since we can then simply neglect the interaction energy.
Thus, we will use this as a common estimate for $R$ in terms of the energy.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbhp]
\centering
\includegraphics[scale=0.65]{ZhiyuPictures/N_5_GS_pre_2_rev-crop.pdf}
\caption{The ground state of a system with $N=5$ particles, shown via the positions of each
  particle, labeled 1 through 5.  At large interactions, the particles position themselves just
  outside the range of interactions of the neighboring particles.}
\label{fig:GS1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

Because of the discontinuous `Heaviside theta' form of our interaction, at large enough interactions
(large $F_0$) the particles are spaced exactly at distance equal to the range $\sigma$, which is
distance $1$ in our rescaled units.  The interaction is then effectively a `hard-core' interaction,
or alternatively, the particles can be thought of as hard rods of length $2\sigma$.

@@

The initial state is taken to have particles with zero velocity and random positions; hence the line
distribution in the phase space picture.  This has the advantage that the breathing motion is
prominently visible.  In addition, the question of long-time relaxation has the simple
interpretation of evolving from the line distribution to a circularly symmetric distribution in
phase space.

@@

We are interested in the oscillations of the size $R(t)$ of the cloud.  It is convenient that our
finite-size simulations start with a line distribution in phase space.  If we started with a state
whose configuration deviates only slightly from circular symmetry in $X$-$P$ space, the oscillating
amplitude would be too small to be distinguished from noise.

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbph]
\includegraphics[width=0.95\textwidth]{ZhiyuPictures/phase-space_collisioncartoons_01.pdf}
\caption{Schematic diagrams of two-particle collision process in the $X_{\rm rot}$-$P_{\rm rot}$
  plane. 
%
The black filled dots shown locations of the two particles in $X_{\rm rot}$-$P_{\rm rot}$ space
before the collision; the empty dots show locations after the collision.  The dashed lines show the
trajectories during the collision process.  Collistions for different $F_0$ values are shown with
different color.
%
\textbf{(a)} When $F_0$ is large, particles cannot pass each other: during the collision the
momentum changes sign.  Being subject to a constant repulsive force, their trajectories during
collision are curved arcs.  The weaker $F_0$ is, the longer the collision lasts, and hence the more
curved is the corresponding arc.
%
For very strong interactions (red dashed line), particles exchange momenta instantaneously; hence
the red dashed line is straight.  The locations of the particles after this collision are on the
gray line; the angle $\alpha$ made by this gray line is estimated in the text.  The yellow, green,
and blue dashed curves are for successively weaker interactions.  \textbf{(b)} When $F_0$ is small,
particles pass each other, so that the force they feel is discontinuous (changes sign) at the moment
when they cross each other.  Their trajectory in $X_{\rm rot}$-$P_{\rm rot}$ space has a
corresponding turning point.  The dashed bright blue trajectory is identical with the one in (a),
which is the critical situation between passing and bouncing. In that case, particles have zero
relative velocity when they collide. }
\label{fig:Breathingfrequency3}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

To analyze the rotation relative to the $X_{\rm rot}$-$P_{\rm rot}$ frame, we consider two-particle
collisions.  In Figure \ref{fig:Breathingfrequency3}(a), we show schematics of such collisions near
the trap center at strong interactions.  
%
During the collision, the particle momentum is changing at a constant rate (the force is a constant
repulsion), so the particle trajectory has a component of motion in the direction of the $P$ axis.
However, the $P$ axis itself is rotating in the $X_{\rm rot}$-$P_{\rm rot}$ frame, at frequency
$\omega_0$.  Thus, the trajectories of the particles during the collision in the $X_{\rm
  rot}$-$P_{\rm rot}$ plane are curved arcs.
%
As long as $F_0$ is large enough that the particles bounce off each other, the particle momentum
will change sign, so a trajectory starting at negative $P$ will end at positive $P$.
%
For very large $F_0$, the particles simply exchange momentum when they are at distance $\sigma$ from
each other; this process is shown in red as two straight lines.  (The particles are initially at
$X=\pm\frac{1}{2}\sigma$ and remain at these $X$ values, but exchange their momenta.)  For smaller
$F_0$, there is change of both position and momentum, as shown in yellow, green and blue for
successively weaker interactions.

The blue dashed line in Figure \ref{fig:Breathingfrequency3}(a) shows a case where $F_0$ is small
enough such that the particles can just cross each other.
%
In Figure \ref{fig:Breathingfrequency3}(b) we show collisions for even smaller $F_0$, so that the
particles pass each other.  The force changes direction discontinuously when the particles cross.
In the $X_{\rm rot}$-$P_{\rm rot}$ plane, this is seen as a sharp turning point in the trajectory
--- if the momentum is initially positive, it first decreases and then increases again.

The boundary case between bouncing and passing behaviors is shown as the blue dashed curve in both
\ref{fig:Breathingfrequency3}(a) and \ref{fig:Breathingfrequency3}(b).  In this case, the momentum
decreases to just about zero when the relative distance reaches zero, so that the particles just
manage to cross.

@@

The precession angle is $\alpha\sim \sigma/P$.  In our rescaled units, the typical momenta of
interacting particles are of the order $R$.  (In phase space $R$ can be interpreted either as the
extent in real space when the velocities are small, or as the extent in momentum space when the
particles approach $x=0$, i.e. when the distribution is along the $P$ direction.)  Thus, we can
estimate $\alpha\sim \sigma/R$, i.e., $\alpha\sim 1/R$, since our unit of length is $\sigma$.  This
is the rotation per unit collision.  Since there are $\sim N$ collistions per period $T=2\pi$, we
have the estimate for the interaction-induced rotation per unit time as
\begin{equation}
\delta ~=~ 2 \frac{N}{2\pi} \frac{1}{R} ~=~  \frac{1}{\pi}  N^{3/2}E^{-1/2}
\label{eq:breathingfrequency1}
\end{equation}
The factor $2$ accounts for the fact that each rotation of the elongated cloud in phase space
corresponds to two breathing mode periods.  In the last step we have used the estimate $E\sim
NR^2$.
